Complete Real-Time Liver Model including Glisson Capsule, Vascular Network and Parenchyma

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
CHANGE: small modifications in Local Deformations section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@545 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{Introduction} 
\IP{I think we should provide some introductory motivation and broader context of the reasarch. I tried this:}

According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
Although new methods such as radio-frequency ablation or cryo-ablation used in interventional radiology 
represent promising improvements, surgery remains the option that offers the foremost success rate against these pathologies. 
Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
of accurate eligibility criteria for the patient. 
In this context, pre-operational planning becomes a crucial task having a significant impact on the treatment.

@@

\subsection{Glisson's Capsule} %{{{

The thickness of the Glisson's capsule is relatively small. Umale et al.
\cite{Umale2011} reported for porcine liver values in range of 10-20
$\mu$m.
It is not possible to model such thin structure with classical tetrahedral
elements. Modeling this non-homogeneous volumetric object would require a
dense mesh to avoid numerical instabilities and would significantly
violate the speed requirements imposed on medical simulators.
Instead modeling the capsule with two-dimensional elements that abstract from the
thickness in the third dimension seems
as a natural choice. Mechanics provides this FE tool in the form of membranes
and shells. Based on the small thickness (compared to the surface area) we also
assume negligible bending forces and propose a model based on membrane
elements. 
To maintain simplicity of the composite model we choose simple triangular
elements with constant strain (CST).

@@

\section{Experiments} %{{{

To evaluate the model it was implemented in the
SOFA\footnote{www.sofa-framework.org} framework and a set of
numerical simulations was performed. First local deformations were compared to the
results reported in literature to validate the method.

@@

In the simulations we used material properties for a porcine liver as
reported by Umale \cite{Umale2013}. The
parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
elastic modulus 8~MPa and thickness 20~$\mu$m.
\IP{OK, now we must decide what value we are going to use for the parenchyma, so it does not explode but it's somehow reasonable...)}
\TG{Right now I'm using 3500 Pa for parenchyma. I can even use higher
modulus, but not any smaller because we get large deformations and use of
corotational model might become questionable.}

\TG{values for blood vessels}

\IP{we have to also add some information how we got the models -- either we comment briefly in each subsection:
Local deformations: cube meshed in gmsh (graded mesh?), number of elements, etc...
Global deformations: contrast-enhanced CT data obtained via acquisition performed on a female pig; the 3D mesh created by a
semi-automatic segmentation using ITKSnap\footnote{www.itksnap.org}, then meshed using CGAL\footnote{www.cgal.org} \cite{Boltcheva2009} resulting in a mesh having 
NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 
created from the segmentation of hepatic vein as described in~\cite{Peterlik2012}. The structured composed of MMM elements.}
\TG{Sure. Put it here or into the respective subsections? (cube: graded
  mesh created by gmsh with 2648 tetrahedra, 518 triangles for capsule
extracted from tetrahedral mesh.)}

\subsection{Local Deformations} %{{{

During the contact with an instrument (probe, needle insertion, sharp
instruments, \ldots) a specific deformation occurs in the near vicinity of
the instrument. This type of deformation may not necessarily induce the
deformation of the object as a whole and therefore can be considered as
local. Correct material properties are not only important for amount of the
displacement but play also a crucial role in capturing the correct area of
the deformation or its profile near the instrument.
\TG{Does it make sense this way? I can add an image. :)}

One example of such local deformation occurs during the aspiration test
when the response of liver to the negative pressure is measured.
Hollenstein et al. \cite{Hollenstein2006} observed that the Glisson's
capsule has non-negligible influence on response of the tissue and modeling
only the parenchyma leads to overestimation of the deformation. We
reproduced the test with our model and acquired similar results.

The aspiration device consists of a tube with 1 cm in diameter and is able
to control internal pressure in the tube. The test is performed by
attaching the tube to the tissue and measuring the tissue response. We
have set up a simulation in the spirit of this test. 
The simulation is performed with a 15x15x15 mm cube of liver tissue and tube
with 1 cm in diameter. The pressure of 3 kPa is simulated inside the tube.

Figure \ref{fig-aspiration} shows the profiles of cuts in the middle of the
test cube. It can be seen that the
deformation for the model without capsule is much larger. This is in good
agreement with the data obtained by Hollenstein et al.

\begin{figure}[htb]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=3in]{aspiration.pdf} &
    \includegraphics[height=1.5in]{aspiration.jpg}\\
    a) & b)
  \end{tabular}

  \caption{\label{fig-aspiration}Aspiration test: a) displacment profile
  with (blue) and without (black) the capsule; b) simulation scene}
\end{figure}

@@

\subsection{Global Deformations}

\begin{figure}[htb]
  \centering
    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
  \caption{\label{f:forceEvol}Evolution of forces during elastic warping.}
\end{figure}

\begin{figure}[htb]
  \centering
    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
  \caption{\label{f:global}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
\end{figure}

In the previous section we have shown that the composite model of parenchyma and 
capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
first we present a real-time complete model of entire liver built from image data. 
Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
the model undergoes large deformations.

The complete model was constructed using constrast-enhanced CT data acquired on a female pig 
being positioned first in supine and then flank configuration. 
The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap.
Further, an unstructured mesh was obtained from the segmented maps using algorithm~\cite{boltcheva}
available in ITKSnap. The semiautomatic method presented in~\cite{peterlik2012} was 
then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
Finally, the model of the Glisson's capsule was constructed using the elements described in 
section~\ref{} over the surface triangles extracted from the tetrahedral mesh. 
The complete model composed of 6200 tetrahedra, 220 beams and 1820 triangles was then 
used in simulation where external force representing the gravity was applied to the model, 
resulting in large deformations mainly in the area of lobes. The refresh rate of 10\,FPS 
was achieved on a desktop computer with Intel i7 CPU running at 3\,GHz and 16\,GB of memory.

>>>>

\section{Introduction} 
According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
Although new methods such as radio-frequency ablation or cryo-ablation used in interventional radiology 
represent promising improvements, surgery remains the option that offers the foremost success rate against these pathologies. 
Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
of accurate eligibility criteria for the patient. 
In this context, pre-operational planning becomes a crucial task having a significant impact on the treatment.

@@

\subsection{Glisson's Capsule} %{{{
\label{ss:capsuleModel}
The thickness of the Glisson's capsule is relatively small. Umale et al.
\cite{Umale2011} reported for porcine liver values in range of 10-20
$\mu$m.
It is not possible to model such thin structure with classical tetrahedral
elements. Modeling this non-homogeneous volumetric object would require a
dense mesh to avoid numerical instabilities and would significantly
violate the speed requirements imposed on medical simulators.
Instead modeling the capsule with two-dimensional elements that abstract from the
thickness in the third dimension seems
as a natural choice. Mechanics provides this FE tool in the form of membranes
and shells. Based on the small thickness (compared to the surface area) we also
assume negligible bending forces and propose a model based on membrane
elements. 
To maintain simplicity of the composite model we choose simple triangular
elements with constant strain (CST).

@@

\section{Experiments} %{{{
To evaluate the model it was implemented in the
SOFA\footnote{www.sofa-framework.org} framework and a set of
numerical simulations was performed. First local deformations were compared to the
results reported in literature to validate the method.

@@

In the simulations we used material properties reported in literature: for the 
Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2013},
whereas for the vessel wall we applied stiffness $E_v$=0.62\,MPa and $t_v$=200\,$\mu$m~\cite{Peterlik2012}.
Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
with $E_p$=3.5\,kPa.

\begin{figure}[t]
    	\subfigure[]{\includegraphics[height=3.5cm]{aspiration.jpg}}
	\subfigure[]{\includegraphics[width=7cm]{aspiration.pdf}}
  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) displacment profile
  with (blue) and without (black) the capsule.}
\end{figure}

\subsection{Local Deformations} %{{{
During the contact with an instrument (probe, needle insertion, sharp
instruments, \ldots) a specific deformation occurs in the near vicinity of
the instrument. This type of deformation may not necessarily induce the
deformation of the object as a whole and therefore can be considered as
local. Correct material properties are not only important to quantify the
displacement, but also play an important role in capturing the correct area of
the deformation or its profile near the instrument.

A good example of such a local deformation is the aspiration test
when the response of liver to the negative pressure is measured.
In~\cite{Hollenstein2006} it is observed that the Glisson's
capsule has a non-negligible influence on the response of the tissue and modeling
only the parenchyma leads to an overestimation of the deformation.

%We reproduced the test with our model and acquired similar results.
The aspiration device consists of a tube with 1\,cm in diameter and is able
to control internal pressure in the tube. The test is performed by
attaching the tube to the tissue and measuring the tissue response. We
have set up a simulation to reproduce the experiment (see Fig.~\ref{fig-aspiration}a: we meshed a 15$\times$15$\times$15\,mm$^3$ 
cube representing the tissue 
resulting in 2648 tetrahedra. The we attached a tube having diameter d=1\,cm 
and applied a pressure of 3\,kPa inside the tube. The interaction between the tube and 
tissue surface was modelled with friction contact.

In Fig.~\ref{fig-aspiration}b the profiles of cuts in the middle of the
test cube are presented showing significantly larger deformation in the model without capsule. This is in perfect 
agreement with the results reported in~\cite{Hollenstein2006} and it can be concluded that 
our composite model based on coupling the triangular and tetrahedral elements provides an accurate 
model of parenchyma and capsule.

@@

\begin{figure}[t]
  \centering
    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
  \caption{\label{f:forceEvol}Evolution of forces during elastic warping.}
\end{figure}

\begin{figure}[t]
  \centering
    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
  \caption{\label{f:global}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
\end{figure}

\subsection{Global Deformations}
In the previous section we have shown that the composite model of parenchyma and 
capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
first we present a real-time complete model of entire liver built from image data. 
Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
the model undergoes large deformations.

The complete model was constructed using constrast-enhanced CT data acquired on a female pig 
being positioned first in supine and then flank configuration. 
The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
The semi-automatic method presented in~\cite{Peterlik2012} was 
then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
Finally, the model of the Glisson's capsule was constructed using the elements described in 
section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
The complete model composed of 6200 tetrahedra, 220 beams and 1820 triangles was then 
used in simulation where external force representing the gravity was applied to the model, 
resulting in large deformations mainly in the area of lobes. The refresh rate of 10\,FPS 
was achieved on a desktop computer with Intel i7 CPU running at 3\,GHz and 16\,GB of memory.
